Austraits -- a curated plant trait database for the Australian flora

### 2019-12-04 ms.Rmd
added manuscript references
added ms referenced

---
title: "Austraits -- a curated plant trait database for the Australian flora"
csl: assets/nature.csl
editor_options:
  chunk_output_type: console
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
  word_document:
    reference_docx: assets/manuscriptstyle.docx
header-includes:
- \usepackage{inputenc}
- \usepackage[T1]{fontenc}
bibliography: references.bib
---

@@

Species traits are essential metrics for comparing ecological strategies in plants arrayed across environmental space or evolutionary lineages1-4. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function5-7. Traits thereby provide useful indicators of species behaviours in communities and ecosystems, regardless of their taxonomy8,9. Through targeted global initiatives the volume of available trait information for plants has grown rapidly in the last two decades. However, the geographic coverage of trait observations across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions. One such region is Australia; a continent with a flora of c. 24,000 native higher-plant species10. While significant investment has been made in curating and digitising herbarium collections in Australia over the last two decades (e.g. The Australian Virtual Herbarium houses ~7 million specimen occurrence records; https://avh.ala.org.au), no complementary resource yet exists for consolidating information on plant traits.

Here we introduce the AusTraits database (hereafter simply AusTraits) - a compilation of plant traits for the Australian flora. Currently, AusTraits draws together 148 primary sources – including published scientific papers, floras, and taxonomic species descriptions– and contains 325,819 measurements spread across 86 different traits for 20,321 species (Table 1; available online only). By providing a harmonised and curated dataset on 86 plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives (but see 11). Global initiatives, such as the TRY database12 and Open Traits Network, have been working towards sharing of data about plant traits though large numbers of Australian plant species remain unrepresented.

@@

  - A reproducible and extensible compilation process from raw data to assembled dataset;
  - Transparent mapping of original values (species names, trait names, units) into the harmonised dataset;
  - High quality data for end users; 
  - Easy addition of new data and correction of community-reported errors; 
  - Satisfying the FAIR principles13, i.e. standards needed to make data resources Findable, Accessible, Interoperable, and Reusable; and
  - Appropriate recognition of contributors.

@@

### Methods

@@

The schema of AusTraits broadly follows the principles of the established Observation and Measurement Ontology14 in that, where available, trait data are connected to contextual information about the collection site (e.g. location co-ordinates, soil conditions, number of replicates) and information about the methods used to derive measurements (e.g. number of replicates, equipment used where available). The target format contains eight core objects, as described in Table 2. The contents of several of these objects are described further in Tables 3-6. This format was developed to include sufficient information about the data and methods used to collect trait measurements in each source. Where possible, trait measurement was recorded as being one of several possible `value_type` (Table 7), reflecting the type of measurement recorded.

@@

To harmonise each source into the common AusTraits format we applied a reproducible and transparent workflow (Figure 1), written in R15 and using custom code and drawing on functions from the packages `tidyverse`16, `stringr`17, `yaml`18, `rprojroot`, `remake`19, `gitr2`, `knitr`20 and `rmarkdown`21. In this workflow, we performed a series of operations, including reformatting data into standardising format, generating observation id’s for each individual measured, transforming variable names into common terms, transforming data common units, standardising terms for categorial variables, encoding suitable metadata, and flagging data that did not pass quality checks.

@@

•	**Automated tests** which confirm that the species names provided can be mapped onto accepted names on The Plant List, that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of accepted values maintained by the creators. Data that does not pass these tests is moved to a separate spreadsheet (“excluded_data”) that is also made available for use if desired.

•	**Detailed review of each source based on bespoke reports**: A report is generated for each source, showing all data and metadata, and is checked by both an AusTraits curator and, where possible, the original contributor for accuracy. In these reports, observations for each trait are plotted against all other values for the trait in AusTraits. This allows AusTraits curators and contributors to quickly visually check the data to ensure it plots against other trait values as expected. Corrections suggested by contributors are combined back into AusTraits and made available with the next static release.

•	**User feedback**: Data issues can be reported for any observation on the project website.

@@

tmp <- RefManageR::WriteBib(austraits$sources, "references.bib", biblatex = F)

@@

# Overview

@@

# Data structures

# Figures

@@

@@

# Tables

>>>>

---
title: "Austraits -- a curated plant trait database for the Australian flora"
csl: assets/nature.csl
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document:
    reference_docx: assets/manuscriptstyle.docx
header-includes:
- \usepackage{inputenc}
- \usepackage[T1]{fontenc}
- \usepackage{float}
bibliography: [ms_ref.bib]
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

@@

Species traits are essential metrics for comparing ecological strategies in plants arrayed across environmental space or evolutionary lineages [@diaz2016global; @kunstler2016plant; @zanne2014three; @cornwell2014functional]. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function [@chapin1993evolution; @adler2014functional; @diaz1998plant]. Traits thereby provide useful indicators of species behaviours in communities and ecosystems, regardless of their taxonomy [@violle2007let; @funk2017revisiting]. Through targeted global initiatives the volume of available trait information for plants has grown rapidly in the last two decades. However, the geographic coverage of trait observations across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions. One such region is Australia; a continent with a flora of c. 24,000 native higher-plant species [@council2010australian]. While significant investment has been made in curating and digitising herbarium collections in Australia over the last two decades (e.g. The Australian Virtual Herbarium houses ~7 million specimen occurrence records; https://avh.ala.org.au), no complementary resource yet exists for consolidating information on plant traits.

Here we introduce the AusTraits database (hereafter simply AusTraits) - a compilation of plant traits for the Australian flora. Currently, AusTraits draws together 148 primary sources – including published scientific papers, floras, and taxonomic species descriptions– and contains 325,819 measurements spread across 86 different traits for 20,321 species (Table 1; available online only). By providing a harmonised and curated dataset on 86 plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives (but see @guerin2017opportunities). Global initiatives, such as the TRY database [@kattge2011try] and Open Traits Network, have been working towards sharing of data about plant traits though large numbers of Australian plant species remain unrepresented.

@@

  - A reproducible and extensible compilation process from raw data to assembled dataset;
  - Transparent mapping of original values (species names, trait names, units) into the harmonised dataset;
  - High quality data for end users; 
  - Easy addition of new data and correction of community-reported errors; 
  - Satisfying the FAIR principles [@wilkinson2016fair], i.e. standards needed to make data resources Findable, Accessible, Interoperable, and Reusable; and
  - Appropriate recognition of contributors.

@@

### Methodology

@@

The schema of AusTraits broadly follows the principles of the established Observation and Measurement Ontology [@madin2007ontology] in that, where available, trait data are connected to contextual information about the collection site (e.g. location co-ordinates, soil conditions, number of replicates) and information about the methods used to derive measurements (e.g. number of replicates, equipment used where available). The target format contains eight core objects, as described in Table 2. The contents of several of these objects are described further in Tables 3-6. This format was developed to include sufficient information about the data and methods used to collect trait measurements in each source. Where possible, trait measurement was recorded as being one of several possible `value_type` (Table 7), reflecting the type of measurement recorded.

@@

To harmonise each source into the common AusTraits format we applied a reproducible and transparent workflow (Figure 1), written in R [@team2018r] and using custom code and drawing on functions from the packages `tidyverse` [@wickham2016tidyverse], `stringr` [@wickham2017stringr], `yaml` [@stephens2014yaml], `rprojroot`, `remake` [@fitzjohn2016remake], `gitr2`, `knitr` [@xie1general] and `rmarkdown` [@allaire2015rmarkdown]. In this workflow, we performed a series of operations, including reformatting data into standardising format, generating observation id’s for each individual measured, transforming variable names into common terms, transforming data common units, standardising terms for categorial variables, encoding suitable metadata, and flagging data that did not pass quality checks.

@@

  - **Automated tests** which confirm that the species names provided can be mapped onto accepted names on The Plant List, that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of accepted values maintained by the creators. Data that does not pass these tests is moved to a separate spreadsheet (“excluded_data”) that is also made available for use if desired.

  - **Detailed review of each source based on bespoke reports**: A report is generated for each source, showing all data and metadata, and is checked by both an AusTraits curator and, where possible, the original contributor for accuracy. In these reports, observations for each trait are plotted against all other values for the trait in AusTraits. This allows AusTraits curators and contributors to quickly visually check the data to ensure it plots against other trait values as expected. Corrections suggested by contributors are combined back into AusTraits and made available with the next static release.

  - **User feedback**: Data issues can be reported for any observation on the project website.

@@

tmp <- RefManageR::WriteBib(austraits$sources, "references.bib")

@@

### Overview

@@

### Data structures

\clearpage

## Figures

@@

\clearpage

@@

## Tables
